Contrasting patterns of richness, abundance, and turnover in mountain bumble bees and their floral hosts

### 2022-01-05 ms_1/ms_revision2.rmd
started final revision

---
title: "Contrasting patterns of richness, abundance, and turnover in mountain bumble bees and their floral hosts"

author: 
  - Douglas B. Sponsler^1^*  
  - Fabrice Requier^2^
  - Katharina Kallnik^1^
  - Alice Classen^1^
  - A. Fabienne Maihoff^1^
  - Johanna Sieger^3^
  - Ingolf Steffan-Dewenter^1^

@@

This study also uses novel code. All data and code required to reproduce our analyses are archived on Dryad: https://doi.org/10.5061/dryad.s1rn8pk7f

@@

Environmental gradients generate and maintain biodiversity on Earth. Mountain slopes are among the most pronounced terrestrial environmental gradients, and the elevational structure of species and their interactions can provide unique insight into the processes that govern community assembly and function in mountain ecosystems. We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyzed the elevational patterns at the nested levels of species abundance, species $\beta$-diversity, and interaction $\beta$-diversity. While floral richness exhibited mid-elevation peak, bumble bee richness increased with elevation before leveling off at the highest sites, demonstrating the exceptional adaptation of these bees to cold temperatures and short growing seasons. In terms of abundance, though, bumble bees exhibited divergent species-level responses to elevation, with a clear separation between species preferring low vs. high elevations. Overall interaction $\beta$-diversity was dominated by strong turnover in the floral community, which exhibited well-defined threshold of $\beta$-diversity rate at the tree line ecotone. Interaction $\beta$-diversity increased sharply at the upper extreme of the elevation gradient (1800-2000 m), an interval over which we also saw steep decline in floral richness and abundance. Turnover of bumble bees along the elevation gradient was modest, with the highest rate of $\beta$-diversity occurring over the interval from low- to mid-elevation sites. The contrast between the relative robustness bumble bees and sensitivity of plants to the elevational gradient in our study suggests that the strongest effects of climate change on mountain bumble bees will be indirect effects mediated by the responses of their floral hosts. Similarly, ongoing changes in land use are likely to cause succession to forest in resource-rich mid-elevation meadows, which may force bumble bees to seek floral resources in agriculturally intensified lowlands or the sparse highlands above the tree line.

@@

In this paper, we present a study based on ~13,000 interactions between bumble bees (Hymenoptera: *Bombus*) and their floral hosts recorded along a 1400 m elevation gradient, complemented by corresponding surveys of floral abundance. Using this dataset, we investigate elevational responses of richness, abundance, and $\beta$-diversity, both at the level of species (bumble bees and flowering plants, respectively) and at the level of interactions (bumble- bee-flower pairs). Our results reveal nonlinear and, in some cases, contrasting elevational patterns, underscoring the fact that mountain slopes are complex ecological systems where multiple processes interact to produce biological responses. These patterns shed light on the basic ecology of the mountain mutualism between bumble bees and wildflowers, and they suggest how this mutualism might be reshaped by ongoing processes of climate and land use change.

@@

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain and when the air temperature was at least 6°C. We divided each 60 x 60 m study plot into 10 evenly-spaced 60 x 6 m transects. During each visit, the observer walked each transect in a zig-zag pattern for five minutes and recorded all interactions between bumble bees and flowers; thus, each sampling event consisted of a total of 50 minutes of observation. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory according to [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot necessarily be interpreted as a foraging visit. In conjunction with bumble bee observations, we estimated the flower cover of each herbaceous or shrubby plant species within the study plot to the nearest 0.01 m^2^; this was done by exhaustive visual inspection of the study plot rather than timed transect walks. To avoid observer bias, floral cover estimates were always performed by the same investigator (K. Kallnik). Plant species were identified following @Lauber2007-ha and @Oberndorfer2001-ro.

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) and bumble bee observations varied across years **(Figure 2A-C)**. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient **(Figure 2D)**.

@@

We quantified bumble bee abundance as the total number of recorded floral visits per bumble bee species per site-date. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *Bombus hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for $\beta$-diversity analyses.

@@

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of $\beta$-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and $\beta$-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the $\beta$-diversity of bumble bees, flora, and their (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between $\beta$-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of sampling intensity, sampling days per site was included as a covariate. We also included the geographic distances between sites as a covariate to account for potential spatial autocorrelation. The importance of the variables in each model were assessed by a permutation test [@Fitzpatrick-2021aa].

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then leveled off and declined slightly at the upper limit of the gradient **(Figure 3A)**. The year 2012 had a negative effect (p = 0.04) on bumble bee richness relative to the reference year 2010 **(Figure 3B)**, but neither mowing nor abandonment had a significant effect on bumble bee richness relative to the reference management level of grazing (p > 0.05)  **(Figure 3C; Table 1)** In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3D)**. The years 2011 (p < 0.001) and 2012 (p < 0.03) had a negative effects on floral richness relative to the reference year 2010 **(Figure 3E)**, but again management had no significant effect (p > 0.1) **(Figure 3F; Table 1)**. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3G)**. The effect of year 2011 on interaction richness was positive (p < 0.05) and that of year 2012 negative (p < 0.001) **(Figure 3H)**, but management did not have a significant effect (p > 0.1) **(Figure 3I; Table 1)**.

Cumulative bumble bee abundance did not respond significantly to elevation (p > 0.1) **(Figure 4A)** or management (p > 0.1) **(Figure 4C)**, but the year 2011 had a positive effect (p = 0.03) on bumble bee abundance relative to the reference year of 2010 **(Figure 4B; Table 2)**. Floral abundance, however, peaked around 1000 and exhibited an approximately quadratic decline with increasing elevation (p = 0.003) **(Figure 4D)**. The effect of year 2012 was negative (p = 0.002) relative to the reference year 2010 (p = 0.02) **(Figure 4E)**, and the effect of abandoned meadows on floral abundance was positive compared to the reference level of grazed meadows **(Figure 4F; Table 2)**.

At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation **(Figure 5; Table 3)**. *Bombus pyrenaeus*, *B. mendax*, *B. monticola*, and *B. mucidus* exhibited approximately linear increases with elevation. In contrast, *Bombus psithyrus* and *B. pascuorum* exhibited approximately linear decreases with elevation. *Bombus pratorum* was unique in exhibiting a strongly unimodal abundance pattern with a well-defined peak between 1500-1600 m, approximately at the tree line ecotone, similar to the abundance pattern of *B. pratorum* described by @Iserbyt2008-jn in the Eyne Valley of the Pyrenees. The remaining species did not respond significantly to elevation (p > 0.05).

@@

Floral $\beta$-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 3A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most vertically distant sites (elevation difference > 1250 m) differed by more than 85%. Bumble bee $\beta$-diversity was, in comparison to floral $\beta$-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 3A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most vertically distant sites remained less than 50%.

Total interaction $\beta$-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most vertically distant sites **(Figure 3B)**. Partitioning revealed that total interaction $\beta$-diversity was driven primarily by species turnover, which accounted for ~50% of total $\beta$-diversity for sites at similar elevation and ~90% of total $\beta$-diversity for the most vertically distant sites. Species turnover consisted mainly of turnover in the floral community **(Figure 3C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most vertically distant sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total $\beta$-diversity, but its share of total $\beta$-diversity declined to ~10% in the most vertically distant sites **(Figure 3B)**.

Bumble bees, flora, and their interactions exhibited marked differences in their rates of $\beta$-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral $\beta$-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 4)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee $\beta$-diversity, 73% of floral $\beta$-diversity, and 71% of interactions $\beta$-diversity.

@@

The roughly sigmoidal pattern of bumble bee $\beta$-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects both the reciprocal abundance patterns of high- and low-elevation species and the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral $\beta$-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium of species composition, but rather a threshold marking the beginning of a process of $\beta$-diversity acceleration that continued for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m [@Korner2003-sc], and the accelerated $\beta$-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance **(Figures 3B, 4B)**.

@@

The poverty of floral resources at the highest elevations raises the question of why bumble bees choose to forage at such elevations at all. For some species, the selection of high elevation habitats may be influenced by thermal tolerances [@Oyen2016-ng], but that fails to explain the presence of elevation-generalist species (including the majority of the species in our study system) in resource-poor high-elevation habitats. It is possible that nest site limitation at lower elevations drives some bumble bees to establish nests at high elevation, and their ability to forage downslope is subsequently constrained by their flight range, which may be reduced due to the energetic costs of flight in cold environments [@Kovac2015-wy]. It is also possible that high elevation landscapes, while sparse in absolute floral resources, offer an escape from more competitive conditions at lower elevations where forager density is higher [@Hoiss2015-gv; @Classen2020-nh]. Similarly, higher elevations could also offer an escape from density-dependent parasite and pathogen pressure [@Parsche2018-ar]. Finally, it is worth speculating that an affinity for open, high-elevation habitats might be an evolutionary legacy in mountain bumble bees, since the existence of flower-rich meadows in the forest matrix below the tree line is largely an artifact of human land use arising mainly in the last 500 years of bumble bees' evolutionary history [@Hejcman2013-ca].

@@

**Figure 2:** Total bumble bee observations per year (A), total samples per year (B), bumble bee observations per sample per year (C), and the distribution of samples through elevation and time (D). Each sample represents a standardized transect walk for a given site and date, each observation represents a recorded interaction between a bumble bee and flower.

**Figure 3:** Conditional effects of elevation, year, and management on the richness of bumble bee species (A), floral species (B), and bumble- bee-flower interactions (C). Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. Bumble bee and plant illustrations were obtained from http://phylopic.org/; see Appendix S1 for details and licensing.

**Figure 4:** Conditional effects of elevation, year, and management on the peak abundance of bumble bees (A) and flower cover (B). Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. The dashed line for the bumble bee abundance smooth indicates that this smooth did not differ significantly from a flat line (p > 0.1).

**Figure 5:** Conditional effects of elevation on the peak abundance of each bumble bee species. Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. Confidence bands are colored to indicated the linear slope of the response curve, and species are plotted in order of decreasing slope.

**Figure 6:** Species-level (A) and interaction-level (B, C) $\beta$-diversity (in terms of proportional turnover) plotted against elevational difference. Each point represents the $\beta$-diversity (or $\beta$-diversity partition) between a pair of sites, and lines represent the overall relationship between $\beta$-diversity and elevation difference using binomial regression smooths. Interaction $\beta$-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned $\beta$-diversity, ST = $\beta$-diversity due to species turnover, OS =  $\beta$-diversity due to interaction rewiring, ST.h = $\beta$-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = $\beta$-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  $\beta$-diversity due to joint species turnover in higher and lower trophic levels.

**Figure 7:** GDM splines of species- and interaction-level $\beta$-diversity in response to elevation. The maximum height of each spline represents the conditional (i.e. partial) effect of elevation on $\beta$-diversity, and the shape of each spline represents to the rate of species turnover as it varies along the elevation gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Rug plots depict the locations of sites along the elevation axis and are color-coded to indicate whether each site fell above or below the tree line.

@@

```{r fig2, echo = FALSE, out.width="100%"}
knitr::include_graphics("../ms_1/figures/fig2.pdf")
```

@@

```{r fig3, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig3.pdf")
```

**Figure 4**

```{r fig4, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig4.pdf")
```

@@

```{r fig5, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig5.pdf")
```

@@

```{r fig6, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig6.pdf")
```

@@

```{r fig7, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig7.pdf")
```

>>>>

---
title: "Contrasting patterns of richness, abundance, and turnover in mountain bumble bees and their floral hosts"
author: 
  - Douglas B. Sponsler^1^*  
  - Fabrice Requier^2^
  - Katharina Kallnik^1^
  - Alice Classen^1^
  - A. Fabienne Maihoff^1^
  - Johanna Sieger^3^
  - Ingolf Steffan-Dewenter^1^

@@

All data and code required to reproduce our analyses are archived on Dryad: https://doi.org/10.5061/dryad.s1rn8pk7f

@@

Environmental gradients generate and maintain biodiversity on Earth. Mountain slopes are among the most pronounced terrestrial environmental gradients, and the elevational structure of species and their interactions can provide unique insight into the processes that govern community assembly and function in mountain ecosystems. We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyzed elevational patterns at the nested levels of species abundance, species $\beta$-diversity, and interaction $\beta$-diversity. While floral richness exhibited mid-elevation peak, bumble bee richness increased with elevation before leveling off at the highest sites, demonstrating the exceptional adaptation of these bees to cold temperatures and short growing seasons. In terms of abundance, though, bumble bees exhibited divergent species-level responses to elevation, with a clear separation between species preferring low vs. high elevations. Overall interaction $\beta$-diversity was dominated by strong turnover in the floral community, which exhibited well-defined threshold of $\beta$-diversity rate at the tree line ecotone. Interaction $\beta$-diversity increased sharply at the upper extreme of the elevation gradient (1800-2000 m), an interval over which we also saw steep decline in floral richness and abundance. Turnover of bumble bees along the elevation gradient was modest, with the highest rate of $\beta$-diversity occurring over the interval from low- to mid-elevation sites. The contrast between the relative robustness bumble bees and sensitivity of plants to the elevational gradient in our study suggests that the strongest effects of climate change on mountain bumble bees will be indirect effects mediated by the responses of their floral hosts. Similarly, ongoing changes in land use are likely to cause succession to forest in resource-rich mid-elevation meadows, which may force bumble bees to seek floral resources in agriculturally intensified lowlands or the sparse highlands above the tree line.

@@

In this paper, we present a study based on ~13,000 interactions between bumble bees and their floral hosts recorded along a 1400 m elevation gradient, complemented by corresponding surveys of floral abundance. Using this dataset, we investigate elevational responses of richness, abundance, and $\beta$-diversity, both at the level of species (bumble bees and flowering plants, respectively) and at the level of interactions (bumble- bee-flower pairs). Our results reveal nonlinear and, in some cases, contrasting elevational patterns, underscoring the fact that mountain slopes are complex ecological systems where multiple processes interact to produce biological responses. These patterns shed light on the basic ecology of the mountain mutualism between bumble bees and wildflowers, and they suggest how this mutualism might be reshaped by ongoing processes of climate and land use change.

@@

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain and when the air temperature was at least 6°C. We divided each 60 x 60 m study plot into 10 evenly-spaced 60 x 6 m transects. During each visit, the observer walked each transect in a zig-zag pattern for five minutes and recorded all interactions between bumble bees and flowers; thus, each sampling event consisted of a total of 50 minutes of observation. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were collected and stored in individually labeled tubes in the freezer for later identification in the laboratory according to @Amiet1996-jz. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot necessarily be interpreted as a foraging visit. In conjunction with bumble bee observations, we estimated the flower cover of each herbaceous or shrubby plant species within the study plot to the nearest 0.01 m^2^ by exhaustive visual inspection of the study plot. To avoid observer bias, floral cover estimates were always performed by the same investigator (K. Kallnik). Plant species were identified following @Lauber2007-ha and @Oberndorfer2001-ro.

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) and of bumble bee observations varied across years. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix S1 for a detailed description of sampling patterns.

@@

We quantified bumble bee abundance as the total number of recorded floral visits per bumble bee species per site-date. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.

@@

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of $\beta$-diversity, fitting each response with a binomial regression smooth. We then used generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to ask whether the relationship between $\beta$-diversity and elevation exhibits thresholds or other nonlinearities. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. We fit separate GDMs for bumble bees, flora, and their (unpartitioned) interactions, respectively. To control for potential confounding effects of sampling intensity, we included sampling days per site as a covariate, and we included the geographic distances between sites as a covariate to account for potential spatial autocorrelation. The importance of the variables in each model were assessed by a permutation test [@Fitzpatrick-2021aa].

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then leveled off and declined slightly at the upper limit of the gradient **(Figure 2A)**. The year 2012 had a negative effect (p = 0.04) on bumble bee richness relative to the reference year 2010 **(Figure 2B)**, but neither mowing nor abandonment had a significant effect on bumble bee richness relative to the reference management level of grazing (p > 0.05)  **(Figure 2C; Table 1)**. In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 2D)**. The years 2011 (p < 0.001) and 2012 (p < 0.03) had a negative effects on floral richness relative to the reference year 2010 **(Figure 2E)**, but again management had no significant effect (p > 0.1) **(Figure 2F; Table 1)**. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 2G)**. The effect of year 2011 on interaction richness was positive (p < 0.05) and that of year 2012 negative (p < 0.001) **(Figure 2H)**, but management did not have a significant effect (p > 0.1) **(Figure 2I; Table 1)**.

Cumulative bumble bee abundance did not respond significantly to elevation (p > 0.1) **(Figure 3A)** or management (p > 0.1) **(Figure 3C)**, but the year 2011 had a positive effect (p = 0.03) on bumble bee abundance relative to the reference year of 2010 **(Figure 3B; Table 2)**. Floral abundance, however, peaked around 1000 and exhibited an approximately quadratic decline with increasing elevation (p = 0.003) **(Figure 3D)**. The effect of year 2012 was negative (p = 0.002) relative to the reference year 2010 (p = 0.02) **(Figure 3E)**, and the effect of abandoned meadows on floral abundance was positive compared to the reference level of grazed meadows **(Figure 3F; Table 2)**.

At the species level, bumble bees exhibited variable and sometimes divergent responses to elevation **(Figure 4; Table 3)**. *Bombus pyrenaeus*, *B. mendax*, *B. monticola*, and *B. mucidus* exhibited approximately linear increases with elevation. In contrast, *Bombus psithyrus* and *B. pascuorum* exhibited approximately linear decreases with elevation. *Bombus pratorum* was unique in exhibiting a strongly unimodal abundance pattern with a well-defined peak between 1500-1600 m, approximately at the tree line ecotone, similar to the abundance pattern of *B. pratorum* described by @Iserbyt2008-jn in the Eyne Valley of the Pyrenees. The remaining species did not respond significantly to elevation (p > 0.05).

@@

Floral $\beta$-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 5A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most vertically distant sites (elevation difference > 1250 m) differed by more than 85%. Bumble bee $\beta$-diversity was, in comparison to floral $\beta$-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 5A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most vertically distant sites remained less than 50%.

Total interaction $\beta$-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most vertically distant sites **(Figure 5B)**. Partitioning revealed that total interaction $\beta$-diversity was driven primarily by species turnover, which accounted for ~50% of total $\beta$-diversity for sites at similar elevation and ~90% of total $\beta$-diversity for the most vertically distant sites. Species turnover consisted mainly of turnover in the floral community **(Figure 5C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most vertically distant sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total $\beta$-diversity, but its share of total $\beta$-diversity declined to ~10% in the most vertically distant sites **(Figure 5B)**.

Bumble bees, flora, and their interactions exhibited marked differences in their rates of $\beta$-diversity in response to elevation **(Figure 6)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral $\beta$-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 4)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee $\beta$-diversity, 73% of floral $\beta$-diversity, and 71% of interactions $\beta$-diversity.

@@

The roughly sigmoidal pattern of bumble bee $\beta$-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects both the reciprocal abundance patterns of high- and low-elevation species and the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral $\beta$-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium of species composition, but rather a threshold marking the beginning of a process of $\beta$-diversity acceleration that continued for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m [@Korner2003-sc], and the accelerated $\beta$-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance **(Figures 2B, 3B)**.

@@

The paucity of floral resources at the highest elevations raises the question of why bumble bees choose to forage at such elevations at all. For some species, the selection of high elevation habitats may be influenced by thermal tolerances [@Oyen2016-ng], but that fails to explain the presence of elevation-generalist species (including the majority of the species in our study system) in resource-poor high-elevation habitats. It is possible that nest site limitation at lower elevations drives some bumble bees to establish nests at high elevation, and their ability to forage downslope is subsequently constrained by their flight range, which may be reduced due to the energetic costs of flight in cold environments [@Kovac2015-wy]. It is also possible that high elevation landscapes, while sparse in absolute floral resources, offer an escape from more competitive conditions at lower elevations where forager density is higher [@Hoiss2015-gv; @Classen2020-nh]. Similarly, higher elevations could also offer an escape from density-dependent parasite and pathogen pressure [@Parsche2018-ar]. Finally, it is worth speculating that an affinity for open, high-elevation habitats might be an evolutionary legacy in mountain bumble bees, since the existence of flower-rich meadows in the forest matrix below the tree line is largely an artifact of human land use arising in the last 500 years of bumble bees' evolutionary history [@Hejcman2013-ca].

@@

**Figure 2:** Conditional effects of elevation, year, and management on the richness of bumble bee species (A), floral species (B), and bumble- bee-flower interactions (C). Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. Bumble bee and plant illustrations were obtained from http://phylopic.org/; see Appendix S1 for details and licensing.

**Figure 3:** Conditional effects of elevation, year, and management on the peak abundance of bumble bees (A) and flower cover (B). Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. The dashed line for the bumble bee abundance smooth indicates that this smooth did not differ significantly from a flat line (p > 0.1).

**Figure 4:** Conditional effects of elevation on the peak abundance of each bumble bee species. Effects are plotted on the link scale, and shaded bands depict 95% confidence intervals. Confidence bands are colored to indicated the linear slope of the response curve, and species are plotted in order of decreasing slope.

**Figure 5:** Species-level (A) and interaction-level (B, C) $\beta$-diversity (in terms of proportional turnover) plotted against elevational difference. Each point represents the turnover (or component of turnover) between a pair of sites, and lines represent the overall relationship between turnover and elevation difference using binomial regression smooths. Interaction turnover (B, C) is partitioned using Poisot et al.’s (2012) components but with more explicit nomenclature: total = unpartitioned interaction turnover, sp. turnover = interaction turnover due to species turnover, OS =  interaction turnover due to interaction rewiring, ST.h = interaction turnover due to species turnover in the higher trophic level (bumble bees), ST.l = $\beta$-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  $\beta$-diversity due to joint species turnover in higher and lower trophic levels.

**Figure 6:** GDM splines of species- and interaction-level $\beta$-diversity in response to elevation. The maximum height of each spline represents the conditional (i.e. partial) effect of elevation on $\beta$-diversity, and the shape of each spline represents to the rate of species turnover as it varies along the elevation gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Rug plots depict the locations of sites along the elevation axis and are color-coded to indicate whether each site fell above or below the tree line.

@@

```{r fig3, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig2.pdf")
```

@@

```{r fig4, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig3.pdf")
```

@@

```{r fig5, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig4.pdf")
```

@@

```{r fig6, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig5.pdf")
```

@@

```{r fig7, echo = FALSE}
knitr::include_graphics("../ms_1/figures/fig6.pdf")
```
